---
title: "CISC889010 | HW2 | Priscilla Hempel"
output: pdf_document
---

Install all the required packages and their dependencies. You may already have some of these installed, depending on your R Studio environment. 
```{r}
 #install.packages("rmarkdown")
 library("rmarkdown")
 #install.packages("boot")
 library("boot")
 library("Matrix")
 #install.packages("mgcv")
 library("mgcv")
 #install.packages("Rgraphviz")
 library("Rgraphviz")
 #install.packages("igraph")
 library("igraph")
 #install.packages("graph")
 library("graph")
# 
 source("https://bioconductor.org/biocLite.R")
 biocLite("graph", siteRepos = "https://bioconductor.org" )
 biocLite("RBGL", siteRepos = "https://bioconductor.org")
 biocLite("Rgraphviz", siteRepos = "https://bioconductor.org")
 biocLite("igraph", siteRepos = "https://bioconductor.org")
# 
 install.packages("gRbase",repos="https://cran.rstudio.com/bin/windows/contrib/3.4/gRbase_1.8-3.zip")
 library("gRbase")

```
In the following sections I describe each of the functions written for this scipt. The MAIN sections, at the end of this document, is where I used the train and test data five times to generate a "best" graph and report on some of its properties. 



This is a function that creates tables for each node of an input graph. The tables contain the counts of each parent = 0 and 1 for an input data set. The columns naming convention is "Parent label:value of parent". 
```{r}

###obtain parents and create Count Tables (ct)###
count_tables <- function(input_graph, traindata){
  
  dag_graph = input_graph
  td = traindata

  #This loop initializes the CPT and orders the cpts by their label (column number)
  ct = list()
  for (i in 1:6) {
    p <- parents(dag_graph@nodes[i], dag_graph)
    
    #if this is a root node (no parents) then just make a count table for the node
    if (length(p)== 0) {
      ct[[strtoi(dag_graph@nodes[i])]] <- data.frame(matrix(0, nrow = 2, ncol = 1))
      
    } 
    
    #if this node has parents make a table to count the values of the parents 
    else {
      ct[[strtoi(dag_graph@nodes[i])]] <- data.frame(matrix(0, nrow = 2, ncol = 2*length(p)))
      
      #Name of the count table columns after the parents and the parents value (parent:value)
      n = 1
      for (nc in 1:length(p)){
        colnames(ct[[strtoi(dag_graph@nodes[i])]])[n] <- paste(p[nc],":0",sep="")
        colnames(ct[[strtoi(dag_graph@nodes[i])]])[n+1] <- paste(p[nc],":1",sep="")
        
        n = n + 2
      }
    }
  }
  
  #This loop fills the ct with counts
  for (i in 1:6){
    p <- parents(dag_graph@nodes[i], dag_graph)
    
    #populate the CPT
    for (j in 1:nrow(td)){
  
      #if node is 0
      if (td[j,strtoi(dag_graph@nodes[i])] == 0){
        
        #if this is a root node
        if (length(p) == 0){
          #calc prob of this parentless (root) node = 0
          ct[[strtoi(dag_graph@nodes[i])]][1,1] <- ct[[strtoi(dag_graph@nodes[i])]][1,1] + 1
        }
  
        else {
        
          prev0 = 0
          #find the parents
          for (k in 1:length(p)){
            #if parent is 0
            if (td[j,strtoi(p[k])] == 0){
              #use double brackets to indicate dataframe and not object in list
              ct[[strtoi(dag_graph@nodes[i])]][1,(prev0+k)] <- ct[[strtoi(dag_graph@nodes[i])]][1,(prev0+k)] + 1
            }
            
            #if parent is 1
            else if(train[j,strtoi(p[k])] == 1){
              ct[[strtoi(dag_graph@nodes[i])]][1,(prev0+k+1)] <- ct[[strtoi(dag_graph@nodes[i])]][1,(prev0+k+1)] + 1
            }
            
            prev0 = k
          }
        }  
      }
      
      #if node is 1
      else if (train[j,strtoi(dag_graph@nodes[i])] == 1){
        
        #if this is a root node
        if (length(p) == 0){
          #count the prob this parentless (root) node = 1
          ct[[strtoi(dag_graph@nodes[i])]][2,1] <- ct[[strtoi(dag_graph@nodes[i])]][2,1] + 1
        }
  
        else {
        
          prev1 = 0
          #find the parents
          for (k in 1:length(p)){
            #if parent is 0
            if (train[j,strtoi(p[k])] == 0){
              #use double brackets to indicate dataframe and not object in list
              ct[[strtoi(dag_graph@nodes[i])]][2,(prev1+k)] <- ct[[strtoi(dag_graph@nodes[i])]][2,(prev1+k)] + 1
            }
            
            #if parent is 1
            else if(train[j,strtoi(p[k])] == 1){
              ct[[strtoi(dag_graph@nodes[i])]][2,(prev1+k+1)] <- ct[[strtoi(dag_graph@nodes[i])]][2,(prev1+k+1)] + 1
            }
            
            prev1 = k
          }
        }
      
      }
    
    }
  
  }
  
  return(ct)
  
}


```


This is a function that will generate the CPT for every node of a graph from an input list of count tables, retrieved from the above count_table function. 

This function returns a list of CPTs. The order of the list corresponds to the numerical label of each node, so the first CPT is the list belongs to the node labeled "1". 

In each CPT, row 1 corresponds to node=0 and row 2 corresponds to node = 1. The columns correspond to each possible combination of the parent values. For a node with one parent, the columns would correspond to parent = 0 and parent = 1. For a node with two parents (p1p2) the order of the columns per the parent values they represent is: 00, 01, 10, 11 . For a node with three parents (p1p2p3) the oder of the columns per the parent values they represent is 000, 001, 010, 011, 100, 101, 110, 111. 
```{r}

get_cpt<-function(input_count_tables){
  ict <- input_count_tables
  
  cpt <- list()  
  for ( i in 1:length(ict)){
    col_count = ncol(ict[[i]])
    
    #Calculate conditional prob of node = 0 and node = 1, given parents value
    if(col_count == 1){ #No parents (root)
      prob <- data.frame(matrix(0, nrow = 2, ncol=1))
      
      prob[1,1] <- ict[[i]][1,1]/colSums(ict[[i]])[1]
      prob[2,1] <- ict[[i]][2,1]/colSums(ict[[i]])[1]
      
      colnames(prob)<-"Root Node Probability"
      cpt[[i]]<-prob
    }
    else if(col_count == 2){ #one parent
      prob0 <- data.frame(matrix(0, nrow = 2, ncol=2))
      
      for (j in 1:2){
          prob0[1,j] <- ict[[i]][1,j]/colSums(ict[[i]])[j]
          prob0[2,j] <- ict[[i]][2,j]/colSums(ict[[i]])[j]
        
      }
      
      #Fill the cpt matrix for a node
      colnames(prob0)<- c("0","1")
      cpt[[i]]<- prob0
      
    }
    else if(col_count == 4){ #Two parents
      prob0 <- data.frame(matrix(0, nrow = 2, ncol=4))
      
      prob0[1,1] <-  (ict[[i]][1,1]+ict[[i]][1,3])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[3])
      prob0[1,2] <-  (ict[[i]][1,1]+ict[[i]][1,4])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[4])
      prob0[1,3] <-  (ict[[i]][1,2]+ict[[i]][1,3])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[3])
      prob0[1,4] <- (ict[[i]][1,2]+ict[[i]][1,4])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[4])
      
      prob0[2,1] <-  (ict[[i]][2,1]+ict[[i]][2,3])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[3])
      prob0[2,2] <-  (ict[[i]][2,1]+ict[[i]][2,4])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[4])
      prob0[2,3] <-  (ict[[i]][2,2]+ict[[i]][2,3])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[3])
      prob0[2,4] <-  (ict[[i]][2,2]+ict[[i]][2,4])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[4])
      
      #Fill the cpt matrix for a node
      colnames(prob0)<- c("00","01","10","11")
      cpt[[i]]<- prob0
      
    
    }
    else if (col_count == 6){ #3 parents
      prob0 <- data.frame(matrix(0, nrow = 2, ncol=8))
      
      prob0[1,1] <-  (ict[[i]][1,1]+ict[[i]][1,3]+ict[[i]][1,5])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[3]+colSums(ict[[i]])[5])
      prob0[1,2] <-  (ict[[i]][1,1]+ict[[i]][1,3]+ict[[i]][1,6])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[3]+colSums(ict[[i]])[6])
      prob0[1,3] <-  (ict[[i]][1,1]+ict[[i]][1,4]+ict[[i]][1,5])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[4]+colSums(ict[[i]])[5])
      prob0[1,4] <-  (ict[[i]][1,1]+ict[[i]][1,4]+ict[[i]][1,6])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[4]+colSums(ict[[i]])[6])
      prob0[1,5] <-  (ict[[i]][1,2]+ict[[i]][1,3]+ict[[i]][1,5])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[3]+colSums(ict[[i]])[5])
      prob0[1,6] <-  (ict[[i]][1,2]+ict[[i]][1,3]+ict[[i]][1,6])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[3]+colSums(ict[[i]])[6])
      prob0[1,7] <-  (ict[[i]][1,2]+ict[[i]][1,4]+ict[[i]][1,5])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[4]+colSums(ict[[i]])[5])
      prob0[1,8] <-  (ict[[i]][1,2]+ict[[i]][1,4]+ict[[i]][1,6])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[4]+colSums(ict[[i]])[6])
      
      prob0[2,1] <-  (ict[[i]][2,1]+ict[[i]][2,3]+ict[[i]][2,5])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[3]+colSums(ict[[i]])[5])
      prob0[2,2] <-  (ict[[i]][2,1]+ict[[i]][2,3]+ict[[i]][2,6])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[3]+colSums(ict[[i]])[6])
      prob0[2,3] <-  (ict[[i]][2,1]+ict[[i]][2,4]+ict[[i]][2,5])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[4]+colSums(ict[[i]])[5])
      prob0[2,4] <-  (ict[[i]][2,1]+ict[[i]][2,4]+ict[[i]][2,6])/(colSums(ict[[i]])[1]+colSums(ict[[i]])[4]+colSums(ict[[i]])[6])
      prob0[2,5] <-  (ict[[i]][2,2]+ict[[i]][2,3]+ict[[i]][2,5])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[3]+colSums(ict[[i]])[5])
      prob0[2,6] <-  (ict[[i]][2,2]+ict[[i]][2,3]+ict[[i]][2,6])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[3]+colSums(ict[[i]])[6])
      prob0[2,7] <-  (ict[[i]][2,2]+ict[[i]][2,4]+ict[[i]][2,5])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[4]+colSums(ict[[i]])[5])
      prob0[2,8] <-  (ict[[i]][2,2]+ict[[i]][2,4]+ict[[i]][2,6])/(colSums(ict[[i]])[2]+colSums(ict[[i]])[4]+colSums(ict[[i]])[6])
      
      #Fill the cpt matrix for a node
      colnames(prob0)<- c("000", "001", "010", "011", "100", "101", "110", "111")
      cpt[[i]]<- prob0
      
    }
  
  }
  return(cpt)  
  
}


```


This is a function that will calculate the score of a given input graph from a data set.
```{r}
score_dag<-function(input_graph, input_data){
  id = input_data
  ig = input_graph
  
  ig_count_tables = count_tables(ig, id)
  ig_cpt = get_cpt(ig_count_tables)
  
  score = 0
  for (i in 1:nrow(id)){
    rowscore = 1
    for (j in 1:length(ig@nodes)){
      
      current_node = strtoi(ig@nodes[j])
      node_value = id[i,current_node]
      
      p_list = parents(ig@nodes[j], ig)
      
      parentvalues = ""
      
      if (is.null(p_list) == FALSE){
        for (k in 1:length(p_list)){
          
          parentvalues = paste(parentvalues, toString(id[i,strtoi(p_list[k])]), sep = "")
          
        }
        
        node_prob = ig_cpt[[current_node]][node_value+1, parentvalues]
      } 
      
      #This is a root node
      else{
        node_prob = ig_cpt[[current_node]][node_value+1,1]
      }
      rowscore = rowscore * node_prob
    }
    score = score + log(rowscore) 
  }
  
  #score = log(score)*-1
  return(score)
  
}

```


This is a function that will return the "best" graph using greed search.
```{r}

get_best<- function(init_graph, dataset){
  g <- score_dag(init_graph, dataset)
  gbest <- init_graph

  loopcheck=TRUE
  while (loopcheck==TRUE){
    ds = FALSE
    
    rnode<-sample(1:6, 2)
    rand_nodes<-list(c(toString(rnode[1]), toString(rnode[2])))
    rand_nodes_reverse <- list(c(toString(rnode[2]), toString(rnode[1])))
    currentg<-gbest
    
    for (o in 1:3){
      
      #Add Edge
      if(o == 1){
        el<- edgeList(gbest)
        
        #Check if an edges exists
        edgeexist1 = TRUE
        
        for (i in 1:length(el)){
          if (identical(el[[i]],rand_nodes[[1]]) || identical(el[[i]],rand_nodes_reverse[[1]])){
             edgeexist1 = FALSE
          }
        }
        
        #If there is not already an edge, add one
        if (edgeexist1 == TRUE){
          new1 <- addEdge(rand_nodes[[1]][1], rand_nodes[[1]][2], gbest, 1)
          ds = TRUE
          #Check if new edge maitains DAG. Check if the score is improved.
          if (is.DAG(new1) && length(parents(rand_nodes[[1]][2], new1)) <= 3 ){
            ns1 <- score_dag(new1, dataset)
            if (ns1 > g){
              gbest <- new1
            }
          }
        }
      }
      
      #Remove Edge
      else if (o == 2){
        el<- edgeList(gbest)
        
        #check if an edge exists and what direction it is
        edgeexist2 = TRUE
        edgeexist3 = TRUE
        for (i in 1:length(el)){
          if (identical(el[[i]], rand_nodes[[1]])){
            edgeexist2 = FALSE
          }
            
          if (identical(el[[i]], rand_nodes_reverse[[1]])){
            edgeexist3 = FALSE
          }
        }
        
        #Remove an edge if it exists in either direction
        if (edgeexist2==FALSE){
          ds = TRUE
          new2 <- removeEdge(rand_nodes[[1]][1],rand_nodes[[1]][2], gbest)
         
          #Check if still a DAG and if the score is better
          if (is.DAG(new2) && length(parents(rand_nodes[[1]][2], new2)) <= 3 ){
            ns2 <- score_dag(new2, dataset)
            if (ns2 > g){
              gbest <- new2
            }
          }
        }
        
        if(edgeexist3==FALSE){
          ds = TRUE
          new2 <- removeEdge(rand_nodes_reverse[[1]][1],rand_nodes_reverse[[1]][2], gbest)
          
          #Check if still a DAG and if the score is better
          if (is.DAG(new2) && length(parents(rand_nodes_reverse[[1]][2], new2)) <= 3 ){
            ns2 <- score_dag(new2, dataset)
            if (ns2 > g){
              gbest <- new2
            }
          }
        }
      }
        
    
      #Reverse edge direction
      else if (o == 3){
        el<- edgeList(gbest)
        
        #check if an edge exists and what direction it is
        edgeexist4 = TRUE
        edgeexist5 = TRUE
        for (i in 1:length(el)){
          if (identical(el[[i]],rand_nodes[[1]])){
            edgeexist4 = FALSE
          }
            
          if (identical(el[[i]],rand_nodes_reverse[[1]])){
            edgeexist5 = FALSE
          }
        }
        
        #Reverse an edge if it exists in either direction
        if (edgeexist4==FALSE){
          ds = TRUE
          new3 <- removeEdge(rand_nodes[[1]][1],rand_nodes[[1]][2], gbest)
          new3 <- addEdge(rand_nodes[[1]][2], rand_nodes[[1]][1], new3, 1)
          
          #Check if still a DAG and if the score is better
          if (is.DAG(new3) && length(parents(rand_nodes[[1]][1], new3)) <= 3 ){
            ns3 <- score_dag(new3, dataset)
            if (ns3 > g){
              gbest <- new3
            }
          }
        }
        if(edgeexist5==FALSE){
          ds = TRUE
          new3 <- removeEdge(rand_nodes_reverse[[1]][1],rand_nodes_reverse[[1]][2], gbest)
          new3 <- addEdge(rand_nodes_reverse[[1]][2],rand_nodes_reverse[[1]][1], new3, 1)
          
          #Check if still a DAG and if the score is better
          if (is.DAG(new3) && length(parents(rand_nodes_reverse[[1]][1], new3)) <= 3 ){
            ns3 <- score_dag(new3, dataset)
            if (ns3 > g){
              gbest <- new3
            }
          }
        }
      }
    }
  
    
    if (identical(gbest,currentg)==TRUE && ds==TRUE){
      loopcheck = FALSE
    }
    else {
      currentg <- gbest
      g = score_dag(gbest, dataset)
    }
  }
  return(gbest)
}

```


This is a function that returns the prediction accuracy of a given gene within a test data set.
```{r}
accuracy <- function(input_graph, trn_data, tst_data, i_gene){
  ig<-input_graph
  dtest<-tst_data
  dtrain<-trn_data
  gene<-i_gene
  
  ig_count_tables = count_tables(ig, dtrain)
  ig_cpt = get_cpt(ig_count_tables)
  
  corr_pred = 0

  for(i in 1:nrow(dtest)){
    gene_cpt<-ig_cpt[[gene]]
    
    p_list = parents(toString(gene), ig)
    parentvalues = ""
    
    gene_value = dtest[i,gene]
    
    #This gene is not a root node
    if (is.null(p_list) == FALSE){
      for (k in 1:length(p_list)){
        parentvalues = paste(parentvalues, toString(dtest[i,strtoi(p_list[k])]), sep = "")
      }
      
      gene_prob = ig_cpt[[gene]][gene_value+1, parentvalues]
      if(gene_prob > 0.5){
        corr_pred = corr_pred+1
        }
      }
    
    #This is a root node
    else{
      gene_prob = ig_cpt[[gene]][gene_value+1,1]
      if(gene_prob > 0.5){
        corr_pred = corr_pred+1
      }
    }
  }
  
  a<-corr_pred/nrow(dtest)
  return(a)
}

```



######
 MAIN
######

First, load the training and the test data. 

```{r}
train <- read.table("hw2_train")
colnames(train)<-c("g1","g2","g3","g4","g5","g6")

test <- read.table("hw2_test")
colnames(test)<-c("g1","g2","g3","g4","g5","g6")
```

Then create an initital network structure, obtain the optimized network structure using the get_best function and plot the graph. Report the accuracy and CPT for the "best" graph. 
Here we repeat this five times.

```{r}
for (i in 1:5){
  #Create inital graph structure. Limit maximum parents = 3. Weight the chances of a node having more parents higher.
  rd<- random_dag(1:ncol(train), maxpar=3, wgt=5)
  
  best <- get_best(rd,train)
  #table(sapply(vpar(best),length))
  print(paste("This is best graph number", i ))
  plot(best)
  
  best_acc<-accuracy(best,train,test,5)
  print(paste("This is the accuracy for best graph number ", i, ": ",best_acc, sep = ""))
  
  ctbls<-count_tables(best,train)
  best_cpt<-get_cpt(ctbls)
  print(paste("This is the CPT for best graph number", i ))
  print(best_cpt)
  
  
}
```


